Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-05-15 AppendixI.Rmd
Discussion

---
title: "Appendix I"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Eric MARCON
    authsuperscript: 1
  - name: Bruno HERAULT
    authsuperscript: 2
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: INPHB (Institut National Ploytechnique Félix Houphoüet Boigny) 
    address: Yamoussoukro, Ivory Coast
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: http://www.ecofog.gf/spip.php?article47
journalinfo: Hal 00679993
date: "`r format(Sys.time(), '%d %B %Y')`"
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
  \usepackage{tabu}
  \renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
  \hyphenation{bio-di-ver-si-ty sap-lings}
lang: english
output: 
  bookdown::gitbook: default
  bookdown::pdf_book:
    base_format: EcoFoG::article
    keep_tex: TRUE
  bookdown::word_document2: default
---

>>>>

---
title: "Appendix I"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Eric MARCON
    authsuperscript: 1
  - name: Bruno HERAULT
    authsuperscript: 2
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: INPHB (Institut National Ploytechnique Félix Houphoüet Boigny) 
    address: Yamoussoukro, Ivory Coast
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: http://www.ecofog.gf/spip.php?article47
journalinfo: Hal 00679993
date: "`r format(Sys.time(), '%d %B %Y')`"
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
  \usepackage{tabu}
  \renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
  \hyphenation{bio-di-ver-si-ty sap-lings}
lang: english
output: 
  bookdown::pdf_book:
    base_format: EcoFoG::article
    keep_tex: TRUE
  bookdown::gitbook: default
  bookdown::word_document2: default
---

### 2018-05-15 WholePlotTrajectories.Rmd
Discussion

@@

The monitoring of disturbed forest communities confirmed a limited negative impact of disturbance on species richness, as observed on several post logging surveys [@Cannon1998;@Baraloto2012a].
Thirty years after disturbance, the genus richness was restored to initial plot values after high disturbance intensity while it substantially increased after low disturbance intensity, reaching a gain of almost 12 genera for some plots.

Both richness and evenness followed asymptotic trajectories after disturbance, sharply increasing for 15 years before stabilizing at higher values than those of control plots.
An increase of communities evenness stems from a higher homogeneity of species distribution.
After disturbance, communities are made of the old, pre-disturbance survivors and the newly recruited trees: changes in composition and abundance distribution are to be found in the recruitment processes and in the pre-disturbance survivors mortality.
Because the composition of old survivors proved to mirror the initial community [@Herault2018], a specific turnover was expected among recruited trees with enhanced growth and survival of previously infrequent species.
Indeed, the increase in taxonomic diversity was accompanied by an increase of taxonomic dissimilarity with plots initial state and a functional shift towards resource-acquisitive strategies (sharp increase in the SLA, leaf thickness and bark thickness and decrease in wood density, leaf toughness and maximum height) [@Westoby1998;@Wright2004;@Reich2014].
Disturbance then causes a reorganization of the typical high dominance structure of hyperdiverse mature forests after disturbance, benefiting to pioneers and light demanding species. 
Likely, the changes in abiotic environment and competitive pressure favored pioneers which outcompete others in non limiting resources but are excluded in mature forests by long-lived, more resistant and shade tolerant species.

As stated by the IDH, communities dynamics after disturbance relied on species functional strategy and corresponding ability to fill the environmental niches made available by disturbance.
Recruited species then mixed with pre-disturbance ones, from which they differed, and constituted a community all the more diversified that the disturbance was intense [@Molino2001].

## On the recovery of disturbed communities

All diversity and composition trajectories followed similar hump-back trajectories returning towards their initial values.
Communities trajectories therefore illustrated their taxonomic and function resilience.

Because functional traits and functional diversity are the most direct link between biodiversity and ecosystem functioning [@Diaz2005], their resilience meant a consistent recovery of ecosystem processes in the long term [@Guariguata2001].
The resilience of the taxonomic diversity and composition similarly translated  the recovery of communities such as before disturbance.
This resilience meant the maintenance of communities initial differences in composition and supported the idea of their convergence towards a determined state [@Hubbell1999;@Molino2001;@Baraloto2012a].
If all plots followed similar hump-back trajectories, though, some significant differences emerged.
To name but one, the taxonomic evenness of one plot displayed an extensive increase until 30 years after disturbance.
The differences trajectories maxima and time path, along with the fact intrinsinc communities differences are maintained, suggested that the trajectories depend on the initial plots composition or on some abiotic parameters [@Anderson2007].

## Functional redundancy of disturbed ecosystems

Despite the recovery of pre-disturbance state there was a time lag between the trajectories followed by communities taxonomic and functional characteristics.
The trajectory of taxonomic composition and evenness had a longer time-path and remained altered while communities functional composition and diversity had already recovered.

Such delay between functional and taxonomic dynamics was already observed for grasslands [@Tilman1997; @Mouillot2011] and more recently for tropical forests [@Lohbeck2015;@Guariguata2001].
According to the "vegetation quantity effect" [@Grime1998], the functional diversity of communities rely on the dominant species.
Communities functional trajectories were then driven first by the increase in diversity and evenness of dominant species, which restored the pre-disturbance functional characteristics.
Then, the functional trajectories were driven by the recruitment of species resembling the old pre-disturbance community, which reduced the overall diversity.
At that time, the taxonomic composition and diversity remained altered: the still missing species were then the initially infrequent and functionally redundant ones.
The functional redundancy, the functional overlap between species that is typical of the huge biodiversity of tropical forests [@Bellwood2006], was then not restored 30 years after disturbance and this is major to consider as it defines forests' resilience [@Trenbath1999;@Elmqvist2003;@Diaz2005].

Besides the long-term alteration of functional redundancy, there was probably persistent compositional changes favoring disturbance resistant species [@Haddad2008], lianas or epiphytes [@Martin2013] and environmental changes, like in the soils nutrient cycling and compaction [@Olander2005].
These persistent changes highly question forest's resilience [@Chazdon2003a].
New conditions would not only be longer lasting but self-maintained as tied to disturbance regime [@Burslem2000].
Specificaly, this would impair species contingent to undisturbed forests, threatening their maintenance, and run the risk to loose cornerstone species and trigger unexpected ecological consequences [@Jones1994;@Diaz2005;@Gardner2007].

>>>>

## Decoupled recovery of communities taxonomic and functional characteristics

Both communities taxonomic and functional diversity and composition proved resilient, following similar humped-back trajectories with a return towards initial values.
The resilience of communities functional characteristics, the most direct link between biodiversity and ecosystem functioning [@Diaz2005], meant a consistent recovery of ecosystem processes in the long term [@Guariguata2001].
The resilience of communities taxonomic characteristics, in turn, meant the maintenance of their initial differences in composition and structure.
It suggested that communities response to disturbance were somehow constraint to converge towards determined compositions [@Hubbell1999; @Molino2001; @Anderson2007; @Baraloto2012a].

Although both communities taxonomic and functional characteristics proved resilient and followed similar humped-back trajectories, the taxonomic recovery systematically lagged behind the corresponding functional dynamics.
Such delay between functional and taxonomic dynamics has already been observed for grasslands [@Tilman1997; @Mouillot2011] and more recently for tropical forests [@Lohbeck2015;@Guariguata2001].
According to the "vegetation quantity effect" [@Grime1998] functional trajectories rely on the pool of dominant species, which diversity and evenness were enhanced after disturbance, and which rapidly restored the functional diversity and composition.
However, communities evenness remained high so infrequent species still missed to the taxonomic recovery that mechanically lagged behind, all the more so that unrecovered species would be functionally redundant and probably undergo competition processes.

@@

Validating the IDH, communities trajectories confirmed the diversity increase after disturbance through an enhanced growth of otherwise less favored species.
This was however completely clear regarding the functional diversity but much more blurry for communities taxonomic characteristics.

The taxonomic richness was weakly or negatively impacted by intense disturbance, as observed on several post logging surveys [@Cannon1998; @Baraloto2012a], while it substantially increased after low intensity disturbance.
Disturbance also enhanced taxonomic evenness, but its intensity was only weakly and lately correlated to communities eveness (*i.e.* Simpson diversity).
Contrastingly, disturbance intensity consistently predicted the significant increase of communities functional diversity for 30 years.
Disturbance impacts involve specific turnover within communities, either among pre-disturbance survivors or among newly recruited trees.
As the composition of old-growth survivors proved to mirror initial communities [@Herault2018], disturbance would impact trees recruitment through the enhanced growth and survival of previously infrequent species and functional characteristics.
Consistently, disturbance resulted in increasing taxonomic dissimilarity compared to pre-disturbance communities and functional shifts towards resource-acquisitive strategies (sharp increase in the SLA, leaf thickness and bark thickness and decrease in wood density, leaf toughness and maximum height) [@Westoby1998;@Wright2004;@Reich2014].
Disturbance then caused a reorganization of the typical high dominance structure of hyperdiverse mature forests after disturbance, benefiting to pioneers and light demanding species. 
Likely, the changes in abiotic environment and competitive pressure favored pioneers which outcompete others in non limiting resources but are excluded in mature forests by long-lived, more resistant and shade tolerant species.
Therefore, consistently with communities dynamics after disturbance relied on species functional strategy and corresponding ability to fill the environmental niches made available by disturbance.
Recruited species then mixed with pre-disturbance ones, from which they differed, and constituted a community all the more diversified that the disturbance was intense [@Molino2001].

## Functional redundancy of disturbed communities

both the lag between taxonomic and functional recovery and the much better consistency of the IDH when the functional diversity is considered rely on the central role of functional redundancy.
The functional redundancy that is the overlap between species, typical of the huge biodiversity of tropical forests [@Bellwood2006] and determinant for communities resilience [@Trenbath1999; @Elmqvist2003; @Diaz2005], was not recovered 30 years after disturbance.

Disturbance globally reduced the overall functional redundancy of communities but this restored rapidly without consistent rtajectories along time.
The initial functional space corresponding to undistrubed 
Besides the long-term alteration of functional redundancy, there was probably persistent compositional changes favoring disturbance resistant species [@Haddad2008], lianas or epiphytes [@Martin2013] and environmental changes, like in the soils nutrient cycling and compaction [@Olander2005].
These persistent changes highly question forest's resilience [@Chazdon2003a].
New conditions would not only be longer lasting but self-maintained as tied to disturbance regime [@Burslem2000].
Specificaly, this would impair species contingent to undisturbed forests, threatening their maintenance, and run the risk to loose cornerstone species and trigger unexpected ecological consequences [@Jones1994;@Diaz2005;@Gardner2007].

### 2018-05-15 WholePlotTrajectories.tex
Discussion

To counter the remaining taxonomic uncertainty, plots taxonomic
composition and diversity were analysed at the genus level, \emph{i.e.}
referring to the genus of observed or trialed botanical names.
Variations in plots taxonomic and functional composition after
disturbance was visualized by their trajectories in a two-dimensional
ordination space over 30 years. Two NMDS were conducted, either from
taxonomic flora inventories or from plots functional composition based
on the 8 leaf, stem and life history traits. Plots trajectories along
time was reported comparatively to the inventories in 1989, 5 years
after disturbance, which corresponded to first inventory with a
sufficient degree of uncertainty (\textless{}30\% of undetermined
trees). The inventories dissimilarity compared to the reference 1989
inventory was reported using occurrence-based (Jaccard) and
abundance-based (Bray-Curtis) similarity measures. The trajectory of
inventories along time was visualized with the euclidean distance in the
two-dimensional ordination space to the 1989 inventory. The trajectories
of the leaf and stem and life traits were also visualized with the
community weighted means (CWM), representing the average trait value in
a community weighted by relative abundance of the species carrying each
value \citep{Diaz2007, Garnier2004}. To compensate the intrinsinc
difference among plots the trajectories corresponded to the differences
along time with the reference inventory in 1989. Species seed mass
corresponded to 5 classes of increasing mass, seed mass trajectories
were therefore reported as the proportion of each class in the
inventories.

@@

\subsection{Disturbance impacts on communities composition and average
functional traits
values}\label{disturbance-impacts-on-communities-composition-and-average-functional-traits-values}

\subsubsection{Composition trajectories}\label{composition-trajectories}

Over time, 828388 trees and 591 botanical species pertaining to 223
genus and 64 botanical families were recorded. Taxonomic and functional
plots trajectories were examined through the trajecotries in a two
dimensional ordination space of successive inventories from 1989 (5
years after disturbance) to 2015 (31 years after disturbance).
Classifications were performed using either abundance-based Bray-Curtis
(Figure \ref{fig:NMDSplans}) or incidence-based Jaccard dissimilarity,
this last having given similar results (data not shown).

Both taxonomic and functional composition of disturbed plots followed
consistent trajectories over time, translating compositional changes
while control plots remained stable. The addition of functional traits
in the two dimensional space revealed a shift of plots composition
towards more acquisitive functional strategies (disturbed plots changed
from high average WD to high average SLA and chlorophyll content). The
dissimilarity of the successive inventories for disturbed plots compared
to their respective reference inventory in 1989 followed unimodal
trajectories (AppendixI, Figure A2). The maximum dissimilarity of plot
strajectories was positively correlated to the disturbance intensity for
taxonomic (\(\rho_{spearman}^{taxonomic}=0.91\)) as well as functional
composition (\(\rho_{spearman}^{taxonomic}=0.96\)). The time at maximum
dissimilarity for taxonomic composition was reached around 26 years
after disturbance while it was 22 years for functional composition. All
trajectories returned towards the initial composition and translated
cyclic compositional changes as visualized in figure (Figure
\ref{fig:NMDSplans}).

@@

\subsubsection{Traits community weighted
means}\label{traits-community-weighted-means}

@@

Except for leaf cholorphyll content, which continued to increase for
some T3 and T2 plots 30 years after disturbance, all traits and seed
mass proportions displayed a unimodal trajectories. The weighted means
of communities specific maximum height at adult stage (\emph{Hmax}),
leaf toughness (\emph{L\_toughness}) and wood specific gravity
(\emph{WD}) remained significantly lower than their initial value and
than these of the control plots (Figure \ref{fig:CWM}). The weighted
means of bark thickness (\emph{Bark\_thick}) similarly remained
substantilly higher than initially for all disturbed plots while the
specfic leaf area (\emph{SLA}) had almost recovered its initial value
and this of the undisturbed plots at the end of the experiment. Once
again the maximum difference to initial state was correlated to the
disturbance intensity (\(\rho_{spearman}^{L_{thickness}}=0.67\),
\(\rho_{spearman}^{L_{chloro}}=0.45\),
\(\rho_{spearman}^{L_{toughness}}=-0.43\),
\(\rho_{spearman}^{SLA}=0.93\), \(\rho_{spearman}^{WD}=-0.78\),
\(\rho_{spearman}^{Bark-thickness}=0.88\),
\(\rho_{spearman}^{Hmax}=-0.48\)).

@@

\subsection{Trajectories of communities diversity after
disturbance}\label{trajectories-of-communities-diversity-after-disturbance}

\subsubsection{Communities taxonomic and functional
diversity}\label{communities-taxonomic-and-functional-diversity}

Trajectories of Richness, Shannon and Simpson taxonomic diversity were
examined at genus level in relation to the 1989 inventories (5 years
after disturbance) (Figure \ref{fig:DivTaxo}).

{\centering \includegraphics[width=1\linewidth]{WholePlotTrajectories_files/figure-latex/DivTaxo-1}

\caption{Trajectories of the difference to the 1989 inventories (5 years after disturbance) over 30 years after disturbance of plots communities \textbf{(a)} Richness, \textbf{(b)} Shannon and \textbf{(c)} Simpson diversities. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3.}\label{fig:DivTaxo}
\end{figure*}

For undisturbed plots the Richness, Shannon and Simpson diversities
remained stable over the 30 years monitored. After low disturbance
intensity the richness increased, reaching a maximum gain of 14
botanical genera (plot 3 from treatment 2). After intense disturbance
however, the taxonomic richness followed a unimodal decrease with a
return to pre-disturbance values. The taxonomic evenness (Shannon and
Simpson diversities) increased after disturbance, whenever its
intensity. The evenness followed a unimodal trajectory with a just
beginning return towards initial values and a maximum value, reached
around 20 years, positively correlated to the disturbance intensity
(\(\rho_{spearman}^{Shannon}=0.86\), and
\(\rho_{spearman}^{Simpson}=0.89\)). Only two T3 plots (plots 8 and 12)
were still increasing 30 years after disturbance \ref{fig:DivTaxo}),
suggesting a similar be much delayed unimodal trajectory.

Communities functional diversity trajectories were examined through the
Rao diversity based on the 7 leaf, stem and life history traits (the
seed mass was excluded as a qualitative variable). The plot 7 from
treatment 1 that displayed a constantly outlying diversity was removed
from the graphical representation for better readability, see figure A3
in appendix for full graph. The functional diversity of all undisturbed
communities remained stable along the 30 years monitored. For all
disturbed communities the functional diversity followed unimodal
trajectories with a marked return towards the initial state (5 years
after disturbance) with values comparable to those of control plots.

The impact of disturbance was examined more accurately at 10, 20 and 30
years after disturbance through the linear correlation between the
initial percentage of AGB removed and the value of communities Simpson
and Rao diversities (both diversities of order 2).

The functional redundancy was measured as the sum of weighted overlap
among species in communities' functional space. Functional space were
defined in the main two-dimensional space of a PCA analysis based on all
leaf and stem and maximum height traits (see appendix I for PCA
details). Plots functional redundancy compared to the initial value
after logging so that only communities dynamics are followed apart from
the impact of logging itself. Communities functional redundancy remained
stable in control plots but after disturbance the redundancy
trajectories were quite variable (See appendix I). Globally after most
intense disturbance (plots T2 and T3) communities redundancy decreased
at first place and then increased to edge, recover or even exceed the
initial value. Considering the functional redundancy restricted to the
functional space of the initial inventory, though, all disturbed plots
followed similar decreasing humped-shaped trajectories
(@ref(fig:RedFun\_rest)). The maximum redundancy loss observed was
positively correlated with the disturbance intensity (XX spearman to be
measured) and none of the disturbed plots had recovered the functional
redundancy within their initial functional space.

@@

@@

The monitoring of disturbed forest communities confirmed a limited
negative impact of disturbance on species richness, as observed on
several post logging surveys \citep{Cannon1998, Baraloto2012a}. Thirty
years after disturbance, the genus richness was restored to initial plot
values after high disturbance intensity while it substantially increased
after low disturbance intensity, reaching a gain of almost 12 genera for
some plots.

@@

\subsection{On the recovery of disturbed
communities}\label{on-the-recovery-of-disturbed-communities}

All diversity and composition trajectories followed similar hump-back
trajectories returning towards their initial values. Communities
trajectories therefore illustrated their taxonomic and function
resilience.

Because functional traits and functional diversity are the most direct
link between biodiversity and ecosystem functioning \citep{Diaz2005},
their resilience meant a consistent recovery of ecosystem processes in
the long term \citep{Guariguata2001}. The resilience of the taxonomic
diversity and composition similarly translated the recovery of
communities such as before disturbance. This resilience meant the
maintenance of communities initial differences in composition and
supported the idea of their convergence towards a determined state
\citep{Hubbell1999, Molino2001, Baraloto2012a}. If all plots followed
similar hump-back trajectories, though, some significant differences
emerged. To name but one, the taxonomic evenness of one plot displayed
an extensive increase until 30 years after disturbance. The differences
trajectories maxima and time path, along with the fact intrinsinc
communities differences are maintained, suggested that the trajectories
depend on the initial plots composition or on some abiotic parameters
\citep{Anderson2007}.

@@

Despite the recovery of pre-disturbance state there was a time lag
between the trajectories followed by communities taxonomic and
functional characteristics. The trajectory of taxonomic composition and
evenness had a longer time-path and remained altered while communities
functional composition and diversity had already recovered.

Such delay between functional and taxonomic dynamics was already
observed for grasslands \citep{Tilman1997, Mouillot2011} and more
recently for tropical forests \citep{Lohbeck2015, Guariguata2001}.
According to the ``vegetation quantity effect'' \citep{Grime1998}, the
functional diversity of communities rely on the dominant species.
Communities functional trajectories were then driven first by the
increase in diversity and evenness of dominant species, which restored
the pre-disturbance functional characteristics. Then, the functional
trajectories were driven by the recruitment of species resembling the
old pre-disturbance community, which reduced the overall diversity. At
that time, the taxonomic composition and diversity remained altered: the
still missing species were then the initially infrequent and
functionally redundant ones. The functional redundancy, the functional
overlap between species that is typical of the huge biodiversity of
tropical forests \citep{Bellwood2006}, was then not restored 30 years
after disturbance and this is major to consider as it defines forests'
resilience \citep{Trenbath1999, Elmqvist2003, Diaz2005}.

>>>>

To counter the remaining taxonomic uncertainty, plots taxonomic
composition and diversity were analysed at the genus level, \emph{i.e.}
referring to the genus of observed or trialed botanical names.
Variations in plots taxonomic and functional composition after
disturbance was visualized by their trajectories over 30 years in a
two-dimensional ordination space. Two NMDS were conducted, either from
taxonomic flora inventories or from plots functional composition based
on the 7 leaf, stem and life history traits (without seed mass classes).
In both cases the NMDS were performed using occurrence-based (Jaccard)
as well as abundance-based (Bray-Curtis) dissimilarity measures.
Trajectories along time were reported through the euclidean distance of
successive inventories to the reference inventories in 1989, 5 years
after disturbance, when the uncertainty degree did not exceed 30\% of
undetermined trees. The trajectories of the leaf and stem and life
traits were also visualized with the community weighted means (CWM),
representing the average trait value in a community weighted by relative
abundance of the species carrying each value
\citep{Diaz2007, Garnier2004}. To compensate the intrinsinc difference
among plots the trajectories corresponded to the differences along time
with the reference inventory in 1989. Species seed mass corresponded to
5 classes of increasing mass, seed mass trajectories were therefore
reported as the proportion of each class in the inventories.

@@

\subsection{Communities Diversity}\label{communities-diversity}

Communities taxonomic diversity trajectories were examined through the
Richness, Shannon and Simpson diversities at genus level, in relation to
the 1989 inventories (5 years after disturbance) (See annexe I). For
undisturbed plots the Richness, Shannon and Simpson diversities remained
stable over the 30 years monitored. After low disturbance intensity the
richness increased, reaching a maximum gain of 14 botanical genera (plot
3 from treatment 2). After intense disturbance, plots taxonomic richness
followed unimodal trajectories first decrease for around ten years and
then increasing to return to pre-disturbance values. For all disturbed
plots the taxonomic evenness (Shannon and Simpson diversities)
increased, following unimodal trajectories with a just beginning return
towards initial values after 30 years. The maximum evenness, reached
after around 20 years, was positively correlated to the disturbance
intensity (\(\rho_{spearman}^{Shannon}=0.86\), and
\(\rho_{spearman}^{Simpson}=0.89\)). Only two T3 plots (plots 8 and 12)
were still increasing 30 years after disturbance \ref{fig:IDHplot},
suggesting similar but delayed trajectories.

Communities functional diversity trajectories were examined through the
Rao diversity based on the 7 leaf, stem and life history traits (the
seed mass was excluded as a qualitative variable). The plot 7 from
treatment 1 displayed a constantly outlying diversity and was removed
from the graphical representation for better readability (see appendix
for full graphs). The functional diversity of all undisturbed
communities remained stable along the 30 years. For all disturbed
communities the functional diversity followed unimodal trajectories with
a return towards the initial state, comparable to the control plots.

The impact of disturbance was examined specifically through the linear
correlation between the disturbance intensity (intial \%AGB removed) and
the Simpson and Rao diversities (diversities of order 2) at three focus
times after disturbance (10, 20 and 30 years) \ref{fig:IDHplot}. The
Simpson diversity was weakly related to the disturbance intensity
(\(R^2<0.25\)) and only from 20 years after disturbance. The Rao
diversity was more strongly correlated to the disturbance intensity
(\(0.55<R^2<0.75\)) for the three times studied. The impact of
disturbance, translated by the slope of the linear correlation, was the
highest 20 years after disturbance.

\subsection{Communities composition}\label{communities-composition}

\subsubsection{Taxonomic and functional
trajectories}\label{taxonomic-and-functional-trajectories}

In the inventories from 1989 (5 years after disturbance) to 2015 (31
years after disturbance), 828388 trees and 591 botanical species
pertaining to 223 genus and 64 botanical families were recorded. The
trajectories of taxonomic and functional composition were visualized in
a two dimensional ordination space mapping the successive inventories
according to their flora and the corresponding traits. Classifications
were performed using either abundance-based Bray-Curtis (Figure
\ref{fig:NMDSplans}) or incidence-based Jaccard dissimilarity. Both
metrics gave similar results, so only analysis using Bray-Curtis
dissimilarity will be discussed here.

While both taxonomic and functional composition remained stable in
undistrubed communities, after disturbed they followed consistent
trajectories over time that revealed significant compositional changes.
Considering the mapping of functional traits in the same dimensional
space, the post-disturbance trajectories corresponded to shifts towards
more acquisitive functional strategies: disturbed communities changed
from high average WD to high average SLA and chlorophyll content. For
disturbed communities the distance of successive inventories to the
reference inventory in 1989 followed unimodal trajectories (AppendixI,
Figure A2). The maximum of communities dissimilarity to their initial
state was positively correlated to the disturbance intensity for both
taxonomic and functional composition
(\(\rho_{spearman}^{taxonomic}=0.91\) and
\(\rho_{spearman}^{functional}=0.96\) respectively). The time at maximum
dissimilarity was reached around 26 years after disturbance for
taxonomic composition and 22 years for functional composition. All
trajectories followed cyclic composition changes with a return towards
the initial composition (Figure \ref{fig:NMDSplans}).

@@

\subsubsection{Traits community weighted means
(CWM)}\label{traits-community-weighted-means-cwm}

@@

Except for leaf chlorophyll content, which continued to increase for
some T3 and T2 plots, all traits and seed mass proportions followed a
unimodal trajectories and either stabilized or returned towards their
initial values 30 years after disturbance. At that time the weighted
means of communities specific maximum height at adult stage
(\emph{Hmax}), leaf toughness (\emph{L\_toughness}) and wood specific
gravity (\emph{WD}) remained significantly lower than their initial
value and than these of the control plots (Figure \ref{fig:CWM}). The
weighted means of bark thickness (\emph{Bark\_thick}) similarly remained
substantilly higher than initially for all disturbed plots while the
specfic leaf area (\emph{SLA}) had almost recovered its initial value,
similar as those of undisturbed plots. For all traits the maximum
difference to initial state was correlated to the disturbance intensity
(\(\rho_{spearman}^{L_{thickness}}=0.67\),
\(\rho_{spearman}^{L_{chloro}}=0.45\),
\(\rho_{spearman}^{L_{toughness}}=-0.43\),
\(\rho_{spearman}^{SLA}=0.93\), \(\rho_{spearman}^{WD}=-0.78\),
\(\rho_{spearman}^{Bark-thickness}=0.88\),
\(\rho_{spearman}^{Hmax}=-0.48\)).

@@

\subsubsection{Functional redundancy}\label{functional-redundancy}

Communities functional redundancy was measured as the sum of weighted
overlap among species in communities' functional space. Functional
spaces were defined in the main two-dimensional space of a PCA analysis
summarizing the 7 leaf, stem, and maximum height traits (see appendix I
for PCA details). Communities functional redundancy remained stable in
control plots but after disturbance the redundancy trajectories were
quite variable (See appendix I). Globally after most intense disturbance
(plots T2 and T3) communities redundancy decreased at first place before
increasing to edge, recover or exceed the initial value.

Considering the functional redundancy restricted to the functional space
of the initial inventory, all disturbed plots followed similar
decreasing humped-shaped trajectories (@ref(fig:RedFun\_rest)). The
maximum redundancy loss observed was positively correlated with the
disturbance intensity (XX spearman to be measured) and none of the
disturbed plots had recovered the functional redundancy within the
initial functional space.

@@

\subsection{Decoupled recovery of communities taxonomic and functional
characteristics}\label{decoupled-recovery-of-communities-taxonomic-and-functional-characteristics}

Both communities taxonomic and functional diversity and composition
proved resilient, following similar humped-back trajectories with a
return towards initial values. The resilience of communities functional
characteristics, the most direct link between biodiversity and ecosystem
functioning \citep{Diaz2005}, meant a consistent recovery of ecosystem
processes in the long term \citep{Guariguata2001}. The resilience of
communities taxonomic characteristics, in turn, meant the maintenance of
their initial differences in composition and structure. It suggested
that communities response to disturbance were somehow constraint to
converge towards determined compositions
\citep{Hubbell1999, Molino2001, Anderson2007, Baraloto2012a}.

Although both communities taxonomic and functional characteristics
proved resilient and followed similar humped-back trajectories, the
taxonomic recovery systematically lagged behind the corresponding
functional dynamics. Such delay between functional and taxonomic
dynamics has already been observed for grasslands
\citep{Tilman1997, Mouillot2011} and more recently for tropical forests
\citep{Lohbeck2015, Guariguata2001}. According to the ``vegetation
quantity effect'' \citep{Grime1998}, functional trajectories rely on the
pool of dominant species which diversity and evenness were enhanced
after disturbance and which rapidly restored the functional diversity
and composition. However, infrequent species that would further lower
the functional diversity and communities evenness still missed to the
recovery of taxonomic characteristics. Taxonomic recovery then
mechanically lagged behind, all the more so that unrecovered species are
functionally redundant and would probably undergo competitive pressures.

@@

The monitoring of disturbed forest communities confirmed a limited or
negative impact of intense disturbance on species richness, as observed
on several post logging surveys \citep{Cannon1998, Baraloto2012a}, while
. Thirty years after disturbance, the genus richness was restored to
initial plot values after high disturbance intensity while it
substantially increased after low disturbance intensity, reaching a gain
of almost 12 genera for some plots.

@@

@@

The functional redundancy, the functional overlap between species that
is typical of the huge biodiversity of tropical forests
\citep{Bellwood2006}, was then not restored 30 years after disturbance
and this is major to consider as it defines forests' resilience
\citep{Trenbath1999, Elmqvist2003, Diaz2005}.
